---
title: "Flourishing"
subtitle: "A framework to improve lives"
date: "`r Sys.Date()`"
output: 
  #tint::tintHtml
  #pdf_document
  word_document
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE,message = FALSE,warning = FALSE}
library(tint); library(tidyverse)
# invalidate cache when the package version changes
knitr::opts_chunk$set(
	cache.extra = packageVersion("tint"),
	tidy = FALSE
)
options(htmltools.dir.version = FALSE)
```

# Why we do what we do

This framework comes from a simple starting point: The intent of supports and services is to help each person to flourish, to achieve a better life.

This aligns with the MDHHS person-centered planning policy, which begins by stating that:^[MDHHS BHDDA Person-Centered Planning Policy (June 5, 2017), p. 1]

> *The purpose of the community mental health system is to support adults and children... to live successfully in their communities — achieving community inclusion and participation, independence, and productivity [and to] to identify and achieve their personal goals.*

# What do we mean by 'a better life'?

People have been asking themselves what it means to live a good life for thousands of years,^[The philosopher Aristotle defined the highest good of human life as happiness, or flourishing (*eudaimonia*).  cf. *Nicomachean Ethics*] and it is among the most crucial questions for each of us to answer.  Here, we will refer to the characteristics that make up a good life as *quality of life*, or QOL for short, relying primarily on research over the past century to arrive at a common and usable definition.

## What makes a good definition?

If we are going to try to define quality of life, it is important that our definition gets a few things right:^[These considerations are drawn from [Cummins, R. (2005). Moving from the quality of life concept to a theory. JIDR, 49(10), 699-706](https://www.ncbi.nlm.nih.gov/pubmed/16162114).]

1. *Multiple dimensions*.  A good life can only be described using multiple dimensions. These are influenced by personal factors, environmental factors, and the interaction between those factors.
2. *Broad enough for everyone*.  We should want to apply the definition to our own lives, and so should others who are different than us.  The basic characteristics of a good life are the same for all people, regardless of culture, gender, disability, etc. 
3. *Both subjective and objective*. Different people have different priorities.  While we can point to objective facts related to QoL, it necessarily includes the point-of-view of the person who is living their life from day to day.  Note that each dimension of the QoL model may have objectively and subjectively defined indicators.

Taken together, these criteria seek to balance the abstract with the specific to arrive at a definition which is well-rounded and understandable.

## What makes a better life?

Keeping our key requirements in mind, we can draw from the broad reservoir of studies on QoL to date to find frameworks which are multi-dimensional^[See systematic review of HRQoL recommending addition of individual and environmental characteristics:  [Bakas, T., et al. (2012)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3548743/).], cross-culturally relevant,^[See [Schalock, R., et al. (2005).](https://www.researchgate.net/profile/Mian_Wang4/publication/7801771_Cross-Cultural_Study_of_Quality_of_Life_Indicators/links/0deec52df448eac34d000000/Cross-Cultural-Study-of-Quality-of-Life-Indicators.pdf).] and which provide both subjective and objective indicators.

Below is a potential model listing essential dimensions of QoL:

```{r echo=FALSE, message=FALSE, warning=FALSE}

tribble(
  ~Area,                  ~Dimension,                ~`Example Indicators`,
  #--|--|----
  "Independence",         "Personal development",    "Education status, personal skills, ADLs, IADLs",
  "            ",         "Self-determination",      "Choices, autonomy, personal control, goals",
  "Social participation", "Interpersonal relations", "Social networks, activities, relationships",
  "                    ", "Social inclusion",        "Community integration, participation, roles",
  "                    ", "Rights",                  "Human (respect/dignity, equality), Legal",
  "Well-being",           "Emotional well-being",    "Safety, positive experiences, self-concept, stress",
  "          ",           "Physical well-being",     "Health, nutrition, recreation/physical exertion",
  "          ",           "Material well-being",     "Financial status, employment, housing, possessions"
) %>%
knitr::kable(caption = 'QoL Dimensions')

```

## Why is a quality of life framework better than others?

One might ask: *Is quality of life the only potential framework that we could use to measure improvement?*, and the answer is *No*, so it is worth discussing other options and briefly reviewing the attributes of each.  Other options include:

- *Symptom reduction*: Measurement of reduction in symptoms related to specific diagnosable conditions.  Symptom scales such as the PHQ-9, GAD-7 and other tools have commonly been used to measure the impact of treatments on specific conditions, but they are more challenging to use for people with multiple co-occuring conditions (MCC).
- *Improving functional status*: Most currently used assessment tools address functional status, measuring the impact of various conditions on broader life areas in terms of their impact on functional ability.  These are broader than symptom scales, and can detect the impact of various symptoms on a particualr functional domain.
- *Health-related quality of life (HRQoL)*:  HRQoL addresses a subset of QoL domains which are related to perceived physical and mental health.  These models typically exclude non-medical areas such as education or rights.

A QoL approach has the following benefits over the approaches mentioned above:

- *Aspirational*:  Rather than being deficit-based, a QoL approach asks people what they want their lives to be and encourages them to work toward that vision.
- *Inclusive*: Instruments and measures from each of the other areas can be included as a part of the QoL framework, since it is broad enough to include each of these areas, and they each contribute to it.
- *Contextual*: 

# Person-Centered Everything

Despite the central position of *person-centered planning* (PCP) to the Medicaid services system in Michigan, in practice it has often been relegated to the planning meeting itself and the preparation for that meeting: ensuring inclusion of family and friends, personal involvement, etc.  While the act of planning remains a cornerstone of the process, it is only the first step of what is needed to truly achieve one's goals.^[Part of the focus on the meeting is due to the auditing focus on the plan document: an instance of *what is measured, is addressed*.]

## Person-Centered *Planning-Doing-Checking-Acting*

This broader scope is already recognized within the state's policy, which indicates that:^[MDHHS BHDDA Person-Centered Planning Policy (June 5, 2017), p. 1]

> *through PCP, a person is engaged in decision-making, problem solving, monitoring progress, and making needed adjustments to goals and supports and services provided in a timely manner.*

In order to determine the success of an individual plan of support, it is necessary to have a strategy to measure improvement in the person's desired areas of focus.  These measurement-based approaches have been gaining traction in their use across populations.^[[Shalock, et al. (2018)](https://aaidd.org/docs/default-source/sis-docs/changes-in-the-field.pdf?sfvrsn=cd8b3021_0) note "an increased emphasis on... conducting outcome evaluation... to assess the degree to which personal goals, positive changes, or benefits have been achieved" in IDD planning.]

Framing the intent of supports and services as improving personal quality of life through person-centered planning also creates a natural bridge to using well-tested quality improvement processes at the individual level.  Perhaps the most applicable of these is the *Plan-Do-Check-Act* (PDCA) cycle, which involves each element of the broad scope of PCP defined above.  The PDCA cycle has already been successfully incorporated into the supports and treatment planning process for people with varying conditions and needs, from intellectual and developmental disabilities, to mental illness, to physical health concerns.

The table below illustrates the relationship between the PDCA process and a fuller definition of the PCP process:

```{r echo=FALSE, message=FALSE, warning=FALSE}

tribble(
  ~Factor,         ~Plan,                         ~Do,                           ~Check,                    ~Act,
  #--|--|----
  "*Questions*",   "What is your life like?",     "What are you working on?",    "Is life better?",      "What next?",
  "",              "What do you want to pursue?", "How are supports provided?",  "",                     "How to improve approach?",
  "",              "What supports do you need?",  "Where do you live/work?",     "",                     "",
  "*Activities*",  "Identify QoL",                "Work on plan",                "Check QoL",            "Revise plan",
  "",              "Assess needs",                "Coordinate services",         "Reassess needs",       "(Repeat cycle)",
  "",              "Develop plan",                "",                            "",                     "",
  "*Quality area*","Structure",                   "Process",                    "Outcome",               "(Re)structure"
) %>%
knitr::kable(caption = 'Alignment of PDCA and PCP')

```




